---
title: "Assignment 4 Juvenile Snowshoe Hares Report"
author: "Halley McVeigh"
date: "11/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(kableExtra)
library(ggplot2)
library(ggbeeswarm)
library(broom)
library(effsize)

```

### Mini "exploratory report" on juvenile snowshoe hares in Bonanza Creek Experimental Forest

This data is from Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. The data can be found [here](https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14)

## Introduction

This report explores the counts and sizes of juvenile snowshoe hares recorded from 1998 - 2012 as part of the Bonanza Creek LTER (Long Term Ecological Research site).

## Data and Analyses



All analyses are in R version 4.0.2 using RStudio version 1.3.1073. 

## Exploratory Findings

### i. Annual juvenile hare trap counts

```{r}
bonanza_hares <- read_csv(here("data", "bonanza_hares.csv"))


juv_hares <- bonanza_hares %>% # create subset of data for juvenile aged hares
  mutate(date = mdy(date)) %>% # change date format to mdy
  mutate(year = year(date)) %>% # create a column with year
  filter(age == "j") # only include data with age of juvenile


ann_juv_hares <- juv_hares %>% # create a subset of data with the count of juvenile hares per year (all sites included)
  group_by(year) %>% 
  summarize(count = n())

ann_juv_hares_summary <- ann_juv_hares %>% # summary table of max() min() mean() and median() of count
  summarize(max_count = max(count),
            min_count = min(count),
            mean_count = mean(count),
            median_count = median(count))
ann_juv_hares_summary %>% 
  kbl(col.names = c("Max count", "Min count", "Mean count", "Median count"),
          caption = "Table 1. Summary statistics of annual juvenile hare trap counts 1998 - 2012. Data: Kielland et al. 2017.",
      ) %>% 
  kable_styling(full_width = FALSE, "bordered")



```


```{r}
# create a barplot of juvenile hare trap counts per year of the study
ggplot(data = ann_juv_hares) +
  geom_col(aes(x = year, y = count), fill = "coral1", alpha = 0.8) +
  theme_minimal() +
  labs(x = "Year", y = "Count", title = "Annual juvenile hare trap counts 1998 - 2012", caption = "Fig. 1 Data: Kielland et al. 2017.") 
  
```

The maximum count of juvenile snowshoe hares was `r max(ann_juv_hares$count)` and was recorded in 1999 whereas the minimum count of `r min(ann_juv_hares$count)` was recorded in 2010. The mean number of juvenile hares trapped for the years 1998 - 2012 was `r mean(ann_juv_hares$count)` and the median was `r median(ann_juv_hares$count)`. There is a large variability in trap counts from year to year. Generally, over the time period this study was carried out, the count of juvenile hares trapped is decreasing. The maximum count may be reflective of a higher degree of effort (time and trap count) during the 1999 season. Conversely, the years of low trap counts may indicate limited time or energy placed on the study, or lower trap numbers in those years. Moving forward, it would be a more robust study to include the time spent during each year as well as the number of traps included in the study each year to better standardize the data. (Table 1, Figure 1)
 
 
### ii. Visualize juvenile hare weights by sex and site
 
```{r}
juv_weights <- juv_hares %>% 
  select("grid", "sex", "weight", "age") %>% 
  filter(!is.na(weight))


p <- ggplot(data = juv_weights) + 
  geom_beeswarm(aes(x = sex, y = weight, color = sex)) +
  geom_boxplot(aes(x = sex, y = weight), fill = NA) +
  labs(x = "Sex", y = "Weight (g)", title = "Juvenile hare weights", caption = "Fig. 2 Juvenile hare weight (g) by sex and Bonanza Creek sites. Coral (female), aqua (male), and gray (NA = undetermined sex) points indicate individual observations for weight (g) of juvenile snowshoe hares. The box endpoints indicate the 25^th^ and 75^th^ percentile values; the black point and the black line inside the box indicate the mean and median weight (g) value for each sex, respectively. Data: Kielland et al. 2017.") +
  theme_light()
 # scale_x_discrete(labels = c("F", "M", "Undetermined")) +
  #theme(legend.position = "none")

grid.labs <- c("Bonanza Black Spruce", "Bonanza mature", "Bonanza riparian")
names(grid.labs) <- c("bonbs", "bonmat", "bonrip")

p + facet_grid(~ grid, labeller = labeller(grid = grid.labs)) +
                 theme(strip.text = element_text(size = 9.5))
```

* **NA values indicate juvenile snowshoe hares where the sex was not determined.**
 
This boxplot (Figure 2) displays the relationship between juvenile snowshoe hare weight (g) and sex (female, male, and NA = undetermined) across three ecological zones. From left to right, labeled in the headers of facets, are the Bonanza Black Spruce, Bonanza mature, and Bonanza riparian zone. These all exist in the Bonanza Creek Experimental Forest. As seen across these bee swarm boxplots, the female juvenile hares are consistent in their weight (g) distribution and do not differ much between sites, with a higher central tendency in the Bonanza Black Spruce zone. The male juvenile hares are more widely distributed in the mature and riparian zones, with a smaller, narrower distribution of weight (g) values in the Black Spruce ecological habitat. The heaviest juvenile hares overall were males in the riparian zone.  
 
 
 
### iii. Juvenile weight comparison (male & female snowshoe hares)
To further compare mean weights for **juvenile male and female snowshoe hares** included in this study.
 
```{r}
juv_weights_summary <- juv_weights %>% 
  filter(!is.na(weight)) %>% 
  filter(!is.na(sex)) %>% 
  group_by(sex) %>% 
  mutate(sex = recode(sex, f = "Female", m = "Male")) %>% 
  summarize(mean_weight = mean(weight),
            sd_weight = sd(weight),
            sample_size = n())

kbl(juv_weights_summary, 
    caption = "Table 2. Juvenile hare weight comparison summary.Data: Kielland et al. 2017.",
  col.names = c("Sex", "Mean weight (g)", "Standard deviation (g)", "Sample size")) %>% 
  kable_styling(full_width = FALSE, "bordered")
```

```{r, include = FALSE}
# pull vectors of weight for male & female juvenile hares

f_weights <- juv_weights %>% 
  filter(sex == "f") %>% 
  pull(weight)

m_weights <- juv_weights %>% 
  filter(sex == "m") %>% 
  pull(weight)


# plot exploratory histograms & qq-plots to check assumptions
hist(f_weights) # pretty normally distributed
qqnorm(f_weights) # relatively linear
hist(m_weights) # relatively normal distribution, 
qqnorm(m_weights) # bit of a s or dip to it, not fully linear
```

```{r}
# t-test, sample sizes are big enough (both male and female well over n=30 so central limit theory holds) and exploratory graphs of histograms and qq-plots displayed general normality in samples -> proceed with two-sample t-test for means comparison.

weight_ttest <- t.test(f_weights, m_weights) # run t-test

weight_ttest_tidy <- tidy(weight_ttest) # tidy version of t-test

weight_effsize <- cohen.d(f_weights, m_weights) # effect size (Cohen's d)
  
```

As may be expected with mammal species, the male juvenile snowshoe hares of this study were, on average (mean weight), heavier than the female juvenile hares (`r juv_weights_summary[2,2]` $\pm$ `r juv_weights_summary[2,3]` g, *n* = `r juv_weights_summary[2,4]` and `r juv_weights_summary[1,2]` $\pm$ `r juv_weights_summary[1,3]` g, *n* = `r juv_weights_summary[1,4]`, respectively; mean $\pm$ 1 standard deviation)

 
 The actual difference in means
 Effect size (Cohen's *d*)
 The outcome of a two-sample t-test
 

 
 
 
### iv. Relationship between juvenile hare weight & hind foot length
Explore the relationship between juvenile snowshoe hare hind foot length (variable `hindft`, recorded in millimeters), and weight. All juvenile hares are included in this section. 

```{r}
juv_weight_hindft <- juv_hares %>% 
  select("weight", "hindft", "sex") %>% 
  filter(!is.na(weight)) %>% 
  filter(!is.na(hindft))

ggplot(data = juv_weight_hindft) +
  geom_point(aes(y = weight, x = hindft, color = sex)) +
  theme_minimal() +
  labs(y = "Weight (g)", x = "Hind foot length (mm)", title = "Relationship between juvenile hare weight & hind foot length", caption = "Fig. 3 Data: Kielland et al. 2017.")
```

An exploration of linear regression
```{r}

```



## Summary

## Citations

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. 